---
title: "Guide to fjoin"
format: html
vignette: >
  %\VignetteEngine{quarto::html}
  %\VignetteIndexEntry{Guide to fjoin}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    out.width: "100%"  
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

<style>
/* Force wrapping for output blocks after code */
div.sourceCode + pre, pre {
  white-space: pre-wrap !important;
  word-break: break-word !important;
}
</style>

```{r setup, include=FALSE}
options(width = 100) # 80

pkg <- function(name) {
  # Inline HTML with inline style attributes as github strips CSS in <style> blocks
  sprintf('<span style="color:#666666; font-weight:bold;">%s</span>', name)
}
```  
  
## Description

`r pkg("fjoin")` is a data frame join package that runs on `r pkg("data.table")`,
using no-copy conversions and case-specific solutions for very fast performance
on equi and non-equi joins of all styles. It plugs seamlessly into tidyverse
pipelines and general workflows, and has features not found in other
frameworks, including NA-safety by default, on-the-fly column selection,
flexible row-order preservation, multiple-match handling on both sides, and an
indicator column for row origin.

## Installation

Latest version from GitHub :
```r
remotes::install_github("trobx/fjoin")
```

## Features

- **General use**: Works with any data-frame–like inputs (e.g. `data.frame`,
`data.table`, tibble, `sf`, `sf`‑tibble, or `list`) and returns an appropriate
class without requiring manual conversions or `r pkg("data.table")` expertise.

- **Scope**: Inner, left, right, full, semi‑, and anti‑joins with equality and
inequality conditions, plus cross joins.

- **NA-safe joins by default**: Missing values are treated sensibly (no matches on
`NA`) unless overridden via `match.na`.

- **On-the-fly column selection**: The `select` argument lets you choose and order
columns for efficient one-liners.

- **Flexible order**: The `order` argument controls which data frame's row order
is preserved in the result (without expensive sorting).

- **Tag row origin**: The `indicate` option adds a Stata‑style marker column
showing each row’s source.

- **Multiple‑match handling on both sides**: Arguments `mult.x` and `mult.y`
specify how to handle multiple matches on either or both sides, including in
semi‑ and anti‑joins.

- **Transparency and mock joins**: Print the resulting `r pkg("data.table")`
code instead of (or as well as) executing, or run “mock” joins without data that
output template code.

- **High performance**: Uses an advanced internal “cookbook” of `r pkg("data.table")`
join recipes and efficiently casts inputs and outputs for speed and memory-efficiency.

- **Lightweight dependency**: Depends only on `r pkg("data.table")`, a mature,
dependency‑free package with long-term stability.

## API

| **fjoin_* functions**                             | **dtjoin* functions**               |
|:---------------------------------------------------|:------------------------------------|
| _`x`/`y` style_                | _Extended `DT[i]` style_       |
| `fjoin_inner()`, `fjoin_left()`, `fjoin_right()`, `fjoin_full()`      | `dtjoin()`                          |
| `fjoin_left_semi` (alias `fjoin_semi`), `fjoin_right_semi()`    | `dtjoin_semi()`                     |
| `fjoin_left_anti` (alias `fjoin_anti`), `fjoin_right_anti()`    | `dtjoin_anti()`                     |
| `fjoin_cross()`                                            | `dtjoin_cross()`                     |

The `fjoin*` family of functions is recommended for mainstream use. The
`dtjoin*` functions handle object classes and write `r pkg("data.table")` code
via a generalisation of `DT[i]` syntax, which the `fjoin*` wrappers translate 
into a familiar `x`/`y` style .

## Examples

```{r}
library(fjoin)
```

These examples use true joins but they carry over to semi- and anti-joins.

### Example 1: Basic usage and options

This example uses `fjoin_full()`, which nests other cases.

```{r}
# data frames
dfQ <- data.table::fread("
id   quantity                    notes   other_cols
 2          5                       ''          ...
 1          6                       ''          ...
 3          7                       ''          ...
NA          8   'oranges (not listed)'          ...
", quote ="'", data.table = FALSE)

dfP <- data.table::fread("
id       item    price   other_cols
NA     apples       10          ...
 3    bananas       20          ...
 2   cherries       30          ...
 1      dates       40          ...
", data.table = FALSE)
```

##### (1) Basic syntax (note the default `match.na = FALSE`)

```{r}
fjoin_full(dfQ, dfP, on = "id")
```

##### (2) Efficient join-select in one line

```{r}
fjoin_full(dfQ, dfP, on = "id", select = c("item", "price", "quantity"))
```

Equivalent to the following in `r pkg("dplyr")` (where the first two lines avoid
joining irrelevant columns):
```{r eval = FALSE}
x <- dfQ |> select(id, quantity)
y <- dfP |> select(id, item, price)
full_join(x, y, join_by(id), na.matches = "never") |>
  select(id, item, price, quantity)
```

##### (3) Indicator column for row origin
1 left only, 2 right only, 3 joined from both. In Stata since 1985!
```{r}
fjoin_full(
  dfQ,
  dfP,
  on = "id",
  select = c("item", "price", "quantity"),
  indicate = TRUE
)
```

##### (4) Switch row order from `x`-then-`y` to `y`-then-`x`
```{r}
fjoin_full(
  dfQ,
  dfP,
  on = "id",
  select = c("item", "price", "quantity"),
  indicate = TRUE,
  order = "right"
)
```

##### (5) Display code instead of executing
```{r}
fjoin_full(
  dfQ,
  dfP,
  on = "id",
  select = c("item", "price", "quantity"),
  indicate = TRUE,
  order = "right",
  do = FALSE
)
```

### Example 2: Using `mult.x` and `mult.y` to reduce a M:M inequality join to 1:1 

`r pkg("data.table")` (`mult`) and `r pkg("dplyr")` (`multiple`) have options for
reducing the cardinality on one side of the join from many ("all") to one
("first" or "last"). `r pkg("fjoin")` (`mult.x`, `mult.y`) will do this on
either side of the join, or on both sides. This example (using `fjoin_left()`)
shows an application to temporally ordered data frames of "events" and
"reactions".

```{r}
# data frames
events <- data.table::fread("
event_id event_ts
       1       10
       2       20
       3       40
", data.table = FALSE)

reactions <- data.table::fread("
reaction_id reaction_ts
          1          30
          2          50
          3          60
", data.table = FALSE)
```

##### (1) For each event, all subsequent reactions (M:M)
```{r}
fjoin_left(
  events,
  reactions,
  on = c("event_ts < reaction_ts"),
)
```

##### (2) For each event, the next reaction (1:M)
```{r}
fjoin_left(
  events,
  reactions,
  on = c("event_ts < reaction_ts"),
  mult.x = "first"
)
```

##### (3) For each event, the next reaction, provided there was no intervening event (1:1)
```{r}
fjoin_left(
  events,
  reactions,
  on = c("event_ts < reaction_ts"),
  mult.x = "first",
  mult.y = "last"
)
```

### Example 3: Flow from `fjoin_*` to `dtjoin*` to `r pkg("data.table")` and outputs
An illustration showing:

- two calls to `fjoin_left()` (commented out), differing in the `order` argument
- the resulting calls to `dtjoin()`, plus `show = TRUE`
- the generated `r pkg("data.table")` code and output.

```{r}
# data frames
set.seed(1)
df_x <- data.frame(id_x = 1:3, col_x = paste0("x", 1:3), val = runif(3))
df_y <- data.frame(id_y = rep(4:2, each = 2), col_y = paste0("y", 1:6), val = runif(6))

# (1) fjoin_left(df_x, df_y, on = "id_x == id_y", mult.x = "first")
dtjoin(
  df_y,
  df_x,
  on = "id_y == id_x",
  mult = "first",
  i.main = TRUE,
  prefix = "R.",
  show = TRUE
)

# (2) fjoin_left(df_x, df_y, on = "id_x == id_y", mult.x = "first", order = "right")
dtjoin(
  df_x,
  df_y,
  on = "id_x == id_y",
  mult.DT = "first",
  nomatch = NULL,
  nomatch.DT = NA,
  prefix = "R.",
  show = TRUE
)
```

## Performance

```{r echo=FALSE}
pal <- c(
  "#B8860B",
  "grey60",
  "#66C2A5",
  "#1B9E77"
)

pl <- function(x) {
  library(ggplot2)
  ggplot(x, aes(x = soln, y = ifelse(is.na(median_secs), 0, median_secs), fill = soln)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = sprintf("%.1f", median_secs)), vjust = -0.3, size = 3) +
    facet_grid(forcats::fct_inorder(style) ~ forcats::fct_inorder(xyargs), switch = "y") +
    scale_fill_manual(values = pal) +
    #facet_wrap(~ interaction(forcats::fct_inorder(position), forcats::fct_inorder(style)), scales = "free_x", ncol = 2) +
    #facet_wrap(~ interaction(forcats::fct_inorder(style), forcats::fct_inorder(position)), scales = "free_x", ncol = 4) +
    scale_y_continuous(limits = c(0, k)) +
    #theme_minimal() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      #strip.placement.y = "left",
      strip.text.y.left = element_text(angle = 0),
    ) +
    labs(
      x = NULL,
      title = x[, unique(description)],
      subtitle = sprintf("median time in secs, %s runs per join",x[, unique(na.omit(N))])
      )
}

dat <- structure(list(style = c("Inner join", "Inner join", "Inner join", 
"Inner join", "Inner join", "Inner join", "Inner join", "Inner join", 
"Left join", "Left join", "Left join", "Left join", "Left join", 
"Left join", "Left join", "Left join", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Full join", "Full join", "Full join", 
"Full join", "Full join", "Full join", "Full join", "Full join"
), xyargs = c("small on left", "small on left", "small on left", 
"small on left", "big on left", "big on left", "big on left", 
"big on left", "small on left", "small on left", "small on left", 
"small on left", "big on left", "big on left", "big on left", 
"big on left", "small on left", "small on left", "small on left", 
"small on left", "big on left", "big on left", "big on left", 
"big on left", "small on left", "small on left", "small on left", 
"small on left", "big on left", "big on left", "big on left", 
"big on left", "small on left", "small on left", "small on left", 
"small on left", "big on left", "big on left", "big on left", 
"big on left"), soln = structure(c(1L, 2L, 3L, 4L, 1L, 2L, 3L, 
4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 
4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 
4L), levels = c("fjoin", "merge", "dplyr", "collapse"), class = "factor"), 
    description = c("No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered"
    ), N = c(5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 
    5L, 5L, 5L, 5L, 5L, 5L, 5L, NA, 5L, 5L, 5L, NA, 5L, NA, NA, 
    5L, 5L, NA, NA, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L), median_secs = c(6.1902577, 
    6.1713217, 19.9907639, 19.7675571, 6.1000215, 6.1203305, 
    13.3964799, 7.0342887, 6.2597127, 6.2432609, 19.9681676, 
    20.1874484, 5.6448957, 5.6632873, 13.6564839, 5.5813086, 
    10.4640203, 10.8960488, 22.8774932, NA, 8.5460365, 12.2154671, 
    14.2776151, NA, 5.9763782, NA, NA, 5.6241829, 5.7766581, 
    NA, NA, 20.3630658, 10.6481448, 10.864006, 23.0229502, 23.5587724, 
    9.8157773, 13.6417876, 15.7693984, 9.0004465)), row.names = c(NA, 
-40L), class = c("data.frame", "data.table"))

```

`r pkg("fjoin")` benchmarks strongly against other frameworks and in particular seems to be more robust to the direction of the join. These examples are very simple, to allow comparison with `merge.data.table` and `collapse` which have limited feature sets, and don't do justice the richer functionality of `r pkg("fjoin")` and `r pkg("dplyr")`. See the [Performance](vignettes/performance.Rmd) article for more detail.
```{r echo=FALSE}
library(data.table)
k <- 30 
setDT(dat)
pl(dat)
```

The performance of `r pkg("fjoin")` straightforwardly reflects `r pkg("data.table")` in the inner and left joins, but the right and full joins perform a bit better than `merge.data.table`. This is typical: `r pkg("fjoin")`'s solutions for join types and additional options that are not straightforward in native `r pkg("data.table")` have been developed with close attention to performance.

## Notes for tidyverse users
Reasons to use `r pkg("fjoin")` are its shorter-and-sweeter syntax, faster large
data performance, and options that `dplyr::*_join()` doesn't support (the
converse is also true --- see below). There's no particular reason to attach it:
you can just plug it into your pipe using a qualified reference with `::`.

```{r message=FALSE}
library(dplyr)

dfQ |>
  as_tibble() |>
  fjoin::fjoin_left(dfP,
                    on = "id",
                    select = c("item", "price", "quantity"),
                    indicate = TRUE,
                    order = "right") |>
  mutate(revenue = price * quantity)
```

It's also worth mentioning that `r pkg("fjoin")` is much more complete than
`r pkg("dtplyr")`'s `r pkg("data.table")`-backed implementation of joins, which
requires object conversions and doesn't respect `r pkg("dplyr")`'s additional
join arguments such as `na.matches` and `multiple` (in fact it silently ignores
them).

On the other hand, `r pkg("fjoin")` currently has no equivalent of
`dplyr::*_join()`'s `relationship` argument: it is silent and laissez-faire
about cardinality. Nor does it yet support rolling joins, which
`r pkg("dplyr")` achieves elegantly via helper functions in `join_by`. These
features will be added. Nested/list‑column joins (supported by `dplyr::*_join()`)
are not planned.

## Notes for `r pkg("data.table")` users
`r pkg("fjoin")` makes `r pkg("data.table")` joins faster to code by providing 
a unified interface that far outclasses `merge.data.table`. It provides one-line 
join operations that produce elaborate `r pkg("data.table")` code that it would
take a great deal of mental effort and typing to do manually. Even in simple
joins, there is little reason *not* to use it: its solutions are as fast 
or faster than `merge`, and even native `r pkg("data.table")` as typically used
(see below).

However, `r pkg("fjoin")` is not a substitute for the full versatility of
`r pkg("data.table")` joins. It does not support:

- compute-on-join, i.e. specifying computations on the joined columns inside
  `j`, rather than simply selecting them (including `r pkg("data.table")`'s
  unique and memory-efficient `by=.EACHI` mechanism for "`j`-as-you-go" computations).
  However, using `do = FALSE` can provide a coding headstart in such cases by
  letting you copy the console output and edit the `j`-expression(s).
- joins by reference of the direct form `DT[i, on = <on> , v := i.v]` (although
  the assignment pattern `DT[, v := fjoin_left(DT, i, on = <on>, select = "v")$v]`
  can be used)
- for now, rolling and overlap joins.

The package actually began life as a tool to solve `r pkg("data.table")`'s
garbling of join columns in non-equi joins. It still does that:

```{r}
library(data.table)

DT1 <- data.table(t=c(5,25,45))
DT2 <- data.table(t_start=c(1,21), t_end=c(10,30))
```

Range join with `r pkg("fjoin")`:
```{r}
dtjoin(DT2, DT1, on=c("t_start <= t", "t_end >= t"), show = TRUE)
```

Compare the output from `r pkg("data.table")`:
```{r message=FALSE}
library(data.table)
DT2[DT1, on=.(t_start <= t, t_end >= t)]
```

The reason `r pkg("fjoin")` prefers `data.frame()` coupled with `setDT()` to the
more usual `list()` in `j` is to avoid an unnecessary deep copy of the joined
columns. Likewise, `r pkg("fjoin")` avoids deep copies on the way in and out by
"shallow-casting" inputs and outputs to and from `data.table`s as necessary.
Shallow conversion is not a safe general way of operating with foreign objects
in `r pkg("data.table")`, but it works for joins, because the input vectors only
need to be read, and the output vectors are guaranteed to be unshared.

One consequence of this is that `r pkg("fjoin")` is faster than `r pkg("data.table")` itself with standard interactive idioms. Consider the following exaggerated example, where the `i` expression is a `data.frame` and we select columns in `j` with `list()` (aliased `.()`) in the usual way:

```{r message=FALSE}
n <- 1e6L; ncolDT <- 2L; ncolDF <- 10L
DT <- data.table(id = rep(1:n, each = 5L), matrix(runif(n * ncolDT), ncol = ncolDT))
DF <- data.frame(id = 1:n, matrix(runif(n * ncolDF), ncol = ncolDF))

bench::mark(
  iterations = 5,
  check      = TRUE,
  data.table = DT[DF, on = .(id), .(id, V1, V2, X1, X3, X5, X7, X9)],
  fjoin      = dtjoin(DT, DF, on = "id", select.i = c("X1", "X3", "X5", "X7", "X9"))
) |> summary() |> subset(select = c("expression", "n_itr", "median",  "mem_alloc"))
```

`r pkg("fjoin")` is faster here because it dodges a call to `as.data.table.data.frame` on the way in, and a call to `as.data.table.list` on the way out, both of which (currently) deep-copy. The `bench::mark` memory measurements [exclude C-level allocations](https://rdatatable-community.github.io/The-Raft/posts/2025-07-14-memory_R-jangorecki/), but the difference between them reflects these R-level copies. This behaviour will eventually change in `r pkg("data.table")`, at which point `r pkg("fjoin")` will revert to the more familiar `j = list()` idiom. 

## Notes for `r pkg("sf")` users

TODO:

- attribute joins between sf objects permitted (not in dplyr)
- sfc-class output columns always detected and refreshed regardless of whether sf
- an example

## Acknowledgments

TODO: data.table team, idesign
