---
title: "Performance check of fjoin against merge.data.table, dplyr, collapse"
format: html
output:
  html_document:
knitr:
  opts_chunk:
    out.width: "100%"   # make figures full-width
---

<style>
/* Force wrapping for output blocks after code */
div.sourceCode + pre, pre {
  white-space: pre-wrap !important;
  word-break: break-word !important;
}
</style>

```{r setup, include=FALSE}
options(width = 100) # console output width 100 characters (default 80) 
pkg <- function(name) {
  # Inline HTML with inline style attributes as github strips CSS in <style> blocks
  sprintf('<span style="color:#666666; font-weight:bold;">%s</span>', name)
}
```  
  
<!-- This article is for the website only and is not part of the package -->
<!-- 
Leave as .Rmd not .qmd because of issues with the use of external data with
quarto vignettes and articles. See e.g. 
https://github.com/r-lib/pkgdown/issues/2846
https://github.com/r-lib/pkgdown/issues/2877
https://github.com/permaverse/flipr/issues/26
-->

This exercise checks the performance of `r pkg("fjoin")` against `r pkg("data.table")` `merge()` (i.e. `merge.data.table`), `r pkg("dplyr")`, and `r pkg("collapse")` on simple equality joins with moderately large laptop-size data. It does not do justice to the much more comprehensive join functionality offered by `r pkg("fjoin")` and `r pkg("dplyr")` compared with the other two.

### Description
The setup consists of two `data.table`s, one "big" (>4GB, 100M rows) and one "small" (100K rows), joined by equality on a single integer column. The test covers the four true joins (inner, left, right, full), swapping the argument positions of the big and small tables. Both tables are randomly ordered in the first set of runs, and keyed in the second. All solutions are tested with like-for-like arguments.

The exercise is then repeated for the case where there are missing values in the join column on each side, which must not be treated as matches (but whose rows must be retained in the result where appropriate). This comparison is limited to `r pkg("fjoin")` and `r pkg("dplyr")`, which are the only two supporting this option.

Details plus version and system information are at the bottom of this note. 

### Results

```{r echo=FALSE}
data_dir <- file.path("fjoin-performance_data") # relative to /vignettes
library(data.table)
```

```{r  echo=FALSE}
store1 <- readRDS(file.path(data_dir, "store1.rds"))
store2 <- readRDS(file.path(data_dir, "store2.rds"))
store3 <- readRDS(file.path(data_dir, "store3.rds"))
store4 <- readRDS(file.path(data_dir, "store4.rds"))

for (d in list(store3,store4)) setnames(d,"position","xyargs",skip_absent=TRUE)
for (d in list(store3,store4)) d[, soln := factor(soln, levels=c("fjoin","merge","dplyr","collapse"))]

k <- 27 # y-axis height
```

```{r echo=FALSE}

pal <- c(
  "#B8860B",
  "grey60",
  "#66C2A5",
  "#1B9E77"
)

pl <- function(x) {
  library(ggplot2)
  ggplot(x, aes(x = soln, y = ifelse(is.na(median_secs), 0, median_secs), fill = soln)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = sprintf("%.1f", median_secs)), vjust = -0.3, size = 3) +
    facet_grid(forcats::fct_inorder(style) ~ forcats::fct_inorder(xyargs), switch = "y") +
    scale_fill_manual(values = pal) +
    #facet_wrap(~ interaction(forcats::fct_inorder(position), forcats::fct_inorder(style)), scales = "free_x", ncol = 2) +
    #facet_wrap(~ interaction(forcats::fct_inorder(style), forcats::fct_inorder(position)), scales = "free_x", ncol = 4) +
    scale_y_continuous(limits = c(0, k)) +
    #theme_minimal() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      #strip.placement.y = "left",
      strip.text.y.left = element_text(angle = 0),
    ) +
    labs(
      x = NULL,
      # title = x[, unique(description)],
      subtitle = sprintf("median time in secs, %s runs per join",x[, unique(na.omit(N))])
      )
}

kt <- function(dat) {
  kableExtra::kbl(dat[, .(total_secs = round(sum(median_secs), 1)), keyby=.(solution=soln)],
                  format="html") |>
    kableExtra::kable_styling(full_width = FALSE)
}

tab <- function(dat) dat[, .(total_secs = round(sum(median_secs), 1)), keyby=.(solution=soln)]
```

##### 1a: No missing values, tables unordered
```{r echo=FALSE}
pl(store1)
kt(store1[!style %ilike% "right join"]) |>
  kableExtra::add_footnote("NB: Total is sum of median times for inner, left, and full joins. Right joins excluded as not comparable.", notation = "none")
```
##### 1b: No missing values, both tables keyed
```{r echo=FALSE}
pl(store2)
kt(store2[!style %ilike% "right join"]) |>
  kableExtra::add_footnote("NB: Total is sum of median times for inner, left, and full joins. Right joins excluded as not comparable.", notation = "none")
```

##### 2a: Missing values (matches disallowed), tables unordered
```{r echo=FALSE}
pl(store3)
kt(store3[!style %ilike% "right join"]) |>
  kableExtra::add_footnote("NB: Total is sum of median times for inner, left, and full joins. Right joins excluded for comparability with the previous scenario.", notation = "none")
```

##### 2b: Missing values (matches disallowed), both tables keyed
```{r echo=FALSE}
pl(store4)
kt(store4[!style %ilike% "right join"]) |>
  kableExtra::add_footnote("NB: Total is sum of median times for inner, left, and full joins. Right joins excluded for comparability with the previous scenario.", notation = "none")
```

#### Main observations

- `r pkg("fjoin")` is the fastest solution overall (while `r pkg("dplyr")` is the least fast), and is never appreciably slower than another solution in any of the ~40 joins considered.

- Not suprisingly, its performance is similar to `merge.data.table()` (where comparison is possible) since both use equivalent and straightforward `r pkg("data.table")` code for inner and left joins. However, it is faster for right and full joins when the cost of the join is large, because it avoids recomputing it to obtain the anti-join of the right table.

- These two `r pkg("data.table")` solutions are much more robust to switching the order of the tables than `r pkg("collapse")` and `r pkg("dplyr")`.

- When the data are ordered, the differences among the solutions are small: they are all fast.

- Disallowing NA matches is very fast (this is the case for `r pkg("dplyr")` also). Comparison with the non-missing joins is because of differences in the size of the join result when missing values are introduced. 

### Setup details
```{r echo=FALSE}
load(file.path(data_dir,"info.RData"))
```
##### Data
```{r echo=FALSE}
as.M <- function(x) {stopifnot(x/1e6==round(x/1e6)); sprintf("%dM", x/1e6)}
as.K <- function(x) {stopifnot(x/1e3==round(x/1e3)); sprintf("%dK", x/1e3)}
as.GB <- function(x) sprintf("%.1fGB", x/1024^3) # actually these are GiB
as.MB <- function(x) sprintf("%.0fMB", x/1024^2) # actually these are MiB
```

- Two tables, each with an integer join column `id` and other columns numeric.
  - `big`: `r as.M(info$nr_big)` rows, `r info$nc_big` columns, size `r as.GB(info$objsize_big)`
  - `small`: `r as.K(info$nr_small)` rows, `r info$nc_small` columns, size `r as.MB(info$objsize_small)`
- The cardinality is M:1 from `big` to `small`. Both tables have `r as.K(info$ndistinct)` distinct values of `id`. In `small` these are unique while in `big` each `id` is present `r as.K(info$nr_big/info$ndistinct)` times.
- In each table `r 100*(1-info$propn_anti)`% of the distinct `id`s (and hence of the rows) have a match in the other table.
- Initially both tables are randomly ordered and there are no missing `id`s in either table. In scenarios with missing `id`s, `r 100*info$propn_NA`% of the values of `id` in each table are `NA`.

##### Solutions
All solutions are tested with like-for-like arguments, suspending any differences in their defaults:

- no cardinality checks (`dplyr::*_join(..., relationship = "many-to-many")`)
- no post-join sorting (`merge(..., sort = FALSE)`)
- no tests for `NA`/`NaN` values (`fjoin::fjoin_*(..., match.na = TRUE)`)
- multiple matches accepted (`collapse::join(..., multiple = TRUE)`)

In the scenario with missing values, the `r pkg("dplyr")` solutions are run with the further argument `na.matches = "never"`, while `r pkg("fjoin")` is run on default arguments, which exclude such matches.

##### Version and system info
```{r echo=FALSE}
cat(
  paste(info$packages,collapse="\n"),
  info$rversion,
  info$sysname,
  info$cpu,
  paste("Cores:", info$cores),
  paste("Threads:", info$threads),
  paste("RAM (GB):", info$ram),
  sep="\n")
```
```{r eval=FALSE, include=FALSE}

LEFT <- data.table(id=1:3, L=paste0("L",1:3))
RIGHT <- data.table(id=4:2, R=paste0("R",1:3))
LEFT
RIGHT

# all joins order by LEFT...
fjoin_left_no_NA_check(LEFT, RIGHT)
merge_left_no_NA_check(LEFT, RIGHT)
dplyr_left_no_NA_check(LEFT, RIGHT)
collapse_left_no_NA_check(LEFT, RIGHT)

fjoin_full_no_NA_check(LEFT, RIGHT)
merge_full_no_NA_check(LEFT, RIGHT)
dplyr_full_no_NA_check(LEFT, RIGHT)
collapse_full_no_NA_check(LEFT, RIGHT)

# ...except for right joins where behaviour differs
fjoin_right_no_NA_check(LEFT, RIGHT)
fjoin_right_order_left_no_NA_check(LEFT, RIGHT) # sorts left (using order="x")
merge_right_no_NA_check(LEFT, RIGHT) # sorts left
dplyr_right_no_NA_check(LEFT, RIGHT) # sorts left
collapse_right_no_NA_check(LEFT, RIGHT)
fjoin_right_no_NA_check(LEFT, RIGHT)
```

