---
title: "Guide to fjoin"
format: html
vignette: >
  %\VignetteEngine{quarto::html}
  %\VignetteIndexEntry{Guide to fjoin}
  %\VignetteEncoding{UTF-8}
# knitr chunk options:
knitr:
  opts_chunk:
    collapse: false      # don't collapse code and output into one block
    out.width: "100%"    # figs 100% width
# execution settings for Quarto:
execute:
  echo: true            # show R code
  warning: false        # suppress warning messages
  message: false        # suppress package startup/diagnostic messages
  cache: false          # disable chunk caching (always render from scratch)
---

<style>
/* 1. Code chunks: wrap long lines and remove scrollbars */
div.sourceCode {
  overflow-x: visible !important;
}
div.sourceCode pre,
pre.sourceCode {
  white-space: pre-wrap !important;
  word-break: break-word !important;
}
/* 2. Output chunks: Quarto-like styling */
.cell-output-stdout pre,
.cell-output-display pre,
.cell-output pre {
  background: white !important;
  border: none !important;
  padding: 0.4em 0 !important;
  white-space: pre-wrap !important;
  word-break: break-word !important;
}
/* 3. Output chunks: remove syntax highlighting, disable link styling (underline) and clicking */
.cell-output-stdout pre *,
.cell-output-display pre *,
.cell-output pre * {
  color: black !important;
  text-decoration: none !important;
  pointer-events: none;
}
</style>

```{r setup, include=FALSE}
options(width = 100) # console output width 100 characters (default 80) 
pkg <- function(name) {
  # Inline HTML with inline style attributes as github strips CSS in <style> blocks
  # style "pkgname" is defined in _pkgdown.yml 
  sprintf('<span class="pkgname">%s</span>', name)
}
```  
  
# Description

`r pkg("fjoin")` is a data frame join package that runs on `r pkg("data.table")`,
providing fast equi and non-equi joins of all styles. It plugs seamlessly into
tidyverse pipelines and general workflows, and has features not found in other
frameworks, including NA-safety by default, on-the-fly column selection,
flexible row-order preservation, multiple-match handling on both sides, and an
indicator column for row origin.

# Installation

Latest version from GitHub :
```r
remotes::install_github("trobx/fjoin")
```

# Features

- **General use**: Works with any data frame-like inputs (e.g. `data.frame`,
`data.table`, tibble, `sf`, `sf`‑tibble, or `list`) and returns an appropriate
class without requiring manual conversions or `r pkg("data.table")` expertise.

- **Scope**: Inner, left, right, full, semi‑, and anti‑joins with equality and
inequality conditions, plus cross joins.

- **NA-safe joins by default**: Missing values are treated sensibly (no matches
on `NA`) unless overridden via `match.na`.

- **On-the-fly column selection**: Choose and order columns with `select` for
efficient one-liners.

- **Flexible order**: The `order` argument controls which data frame's row order
is preserved in the result (without expensive sorting).

- **Row origin tagging**: The `indicate` option adds a Stata‑style marker column
showing each row’s source.

- **Multiple‑match handling on both sides**: Arguments `mult.x` and `mult.y`
specify how to handle multiple matches on either or both sides, including in
semi‑ and anti‑joins.

- **Transparency and mock joins**: Print the resulting `r pkg("data.table")`
code instead of (or as well as) executing it, or run “mock” joins without data
that output template code.

- **High performance**: Uses an advanced internal “cookbook” of `r pkg("data.table")`
join recipes and efficiently casts inputs and outputs for speed and memory-efficiency.

- **Lightweight dependency**: Depends only on `r pkg("data.table")`, a mature,
dependency‑free package with long-term stability.

# API

| **`fjoin_*` functions**                             | **`dtjoin*` functions**               |
|:---------------------------------------------------|:------------------------------------|
| _`x`/`y` style_                | _Extended `DT[i]` style_       |
| `fjoin_inner()`, `fjoin_left()`, `fjoin_right()`, `fjoin_full()`      | `dtjoin()`                          |
| `fjoin_left_semi` (alias `fjoin_semi`), `fjoin_right_semi()`    | `dtjoin_semi()`                     |
| `fjoin_left_anti` (alias `fjoin_anti`), `fjoin_right_anti()`    | `dtjoin_anti()`                     |
| `fjoin_cross()`                                            | `dtjoin_cross()`                     |

The `fjoin_*` functions are the ones to use. The `dtjoin*` functions
handle object classes and write `r pkg("data.table")` code via a generalisation
of `DT[i]` syntax, which the `fjoin_*` wrappers translate into a familiar `x`/`y`
style.

# Examples

```{r}
library(fjoin)
```

## Example 1: Basic usage and options
This example uses `fjoin_full()`, which nests other cases.
```{r}
# data frames
dfQ <- data.table::fread("
id quantity                   notes other_cols
 2        5                      ''        ...
 1        6                      ''        ...
 3        7                      ''        ...
NA        8  'oranges (not listed)'        ...
", quote ="'", data.table = FALSE)

dfP <- data.table::fread("
id     item price other_cols
NA   apples    10        ...
 3  bananas    20        ...
 2 cherries    30        ...
 1    dates    40        ...
", data.table = FALSE)
```

### (1) Basic syntax
Default `match.na = FALSE` avoids apples-and-oranges matches on missing values.
```{r}
fjoin_full(dfQ, dfP, on = "id")
```

### (2) Efficient selective joins in one line
```{r}
fjoin_full(dfQ, dfP, on = "id", select = c("item", "price", "quantity"))
```

Equivalent to the following in `r pkg("dplyr")` (where the first two lines avoid
inflating the join with irrelevant columns):
```{r eval = FALSE}
x <- dfQ |> select(id, quantity)
y <- dfP |> select(id, item, price)
full_join(x, y, join_by(id), na.matches = "never") |>
  select(id, item, price, quantity)
```

### (3) Indicator column for row origin
1: left only, 2: right only, 3: joined from both. In Stata since 1985!
```{r}
fjoin_full(
  dfQ,
  dfP,
  on = "id",
  select = c("item", "price", "quantity"),
  indicate = TRUE
)
```

### (4) Switch row order from left-then-right to right-then-left
```{r}
fjoin_full(
  dfQ,
  dfP,
  on = "id",
  select = c("item", "price", "quantity"),
  indicate = TRUE,
  order = "right"
)
```

### (5) Display code instead of executing
For `r pkg("data.table")` users. 
```{r}
fjoin_full(
  dfQ,
  dfP,
  on = "id",
  select = c("item", "price", "quantity"),
  indicate = TRUE,
  order = "right",
  do = FALSE
)
```

## Example 2: Reducing an inequality join from M:M to 1:1 with `mult.x` and `mult.y`

`r pkg("data.table")` (`mult`) and `r pkg("dplyr")` (`multiple`) can reduce the
cardinality on one side of the join from many ("all") to one ("first" or "last").
`r pkg("fjoin")` (`mult.x`, `mult.y`) will do this on either side of the join,
or on both sides. This example (using `fjoin_left()`) shows an application to 
temporally ordered data frames of generic "events" and "reactions".

```{r}
# data frames
events <- data.table::fread("
event_id event_ts
       1       10
       2       20
       3       40
", data.table = FALSE)

reactions <- data.table::fread("
reaction_id reaction_ts
          1          30
          2          50
          3          60
", data.table = FALSE)
```

### (1) For each event, all subsequent reactions (M:M)
```{r}
fjoin_left(
  events,
  reactions,
  on = c("event_ts < reaction_ts"),
)
```

### (2) For each event, the next reaction (1:M)
```{r}
fjoin_left(
  events,
  reactions,
  on = c("event_ts < reaction_ts"),
  mult.x = "first"
)
```

### (3) For each event, the next reaction, provided there was no intervening event (1:1)
```{r}
fjoin_left(
  events,
  reactions,
  on = c("event_ts < reaction_ts"),
  mult.x = "first",
  mult.y = "last"
)
```

## Example 3: `fjoin_*` to `dtjoin*` to `r pkg("data.table")` and outputs
An illustration showing:

- two calls to `fjoin_left()` (commented out), differing in the `order` argument
- the resulting calls to `dtjoin()`, plus `show = TRUE`
- the generated `r pkg("data.table")` code and output.

```{r}
# data frames
set.seed(1)
df_x <- data.frame(id_x = 1:3, col_x = paste0("x", 1:3), val = runif(3))
df_y <- data.frame(id_y = rep(4:2, each = 2), col_y = paste0("y", 1:6), val = runif(6))

# (1) fjoin_left(df_x, df_y, on = "id_x == id_y", mult.x = "first")
dtjoin(
  df_y,
  df_x,
  on = "id_y == id_x",
  mult = "first",
  i.home = TRUE,
  prefix = "R.",
  show = TRUE
)

# (2) fjoin_left(df_x, df_y, on = "id_x == id_y", mult.x = "first", order = "right")
dtjoin(
  df_x,
  df_y,
  on = "id_x == id_y",
  mult.DT = "first",
  nomatch = NULL,
  nomatch.DT = NA,
  prefix = "R.",
  show = TRUE
)
```

# Performance

```{r echo=FALSE}
pal <- c(
  "#B8860B",
  "grey60",
  "#66C2A5",
  "#1B9E77"
)

pl <- function(x) {
  library(ggplot2)
  ggplot(x, aes(x = soln, y = ifelse(is.na(median_secs), 0, median_secs), fill = soln)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = sprintf("%.1f", median_secs)), vjust = -0.3, size = 3) +
    facet_grid(forcats::fct_inorder(style) ~ forcats::fct_inorder(xyargs), switch = "y") +
    scale_fill_manual(values = pal) +
    #facet_wrap(~ interaction(forcats::fct_inorder(position), forcats::fct_inorder(style)), scales = "free_x", ncol = 2) +
    #facet_wrap(~ interaction(forcats::fct_inorder(style), forcats::fct_inorder(position)), scales = "free_x", ncol = 4) +
    scale_y_continuous(limits = c(0, k)) +
    #theme_minimal() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      #strip.placement.y = "left",
      strip.text.y.left = element_text(angle = 0),
    ) +
    labs(
      x = NULL,
      title = x[, unique(description)],
      subtitle = sprintf("median time in secs, %s runs per join",x[, unique(na.omit(N))])
      )
}

dat <- structure(list(style = c("Inner join", "Inner join", "Inner join", 
"Inner join", "Inner join", "Inner join", "Inner join", "Inner join", 
"Left join", "Left join", "Left join", "Left join", "Left join", 
"Left join", "Left join", "Left join", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Full join", "Full join", "Full join", 
"Full join", "Full join", "Full join", "Full join", "Full join"
), xyargs = c("small on left", "small on left", "small on left", 
"small on left", "big on left", "big on left", "big on left", 
"big on left", "small on left", "small on left", "small on left", 
"small on left", "big on left", "big on left", "big on left", 
"big on left", "small on left", "small on left", "small on left", 
"small on left", "big on left", "big on left", "big on left", 
"big on left", "small on left", "small on left", "small on left", 
"small on left", "big on left", "big on left", "big on left", 
"big on left", "small on left", "small on left", "small on left", 
"small on left", "big on left", "big on left", "big on left", 
"big on left"), soln = structure(c(1L, 2L, 3L, 4L, 1L, 2L, 3L, 
4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 
4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 
4L), levels = c("fjoin", "merge", "dplyr", "collapse"), class = "factor"), 
    description = c("No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered"
    ), N = c(5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 
    5L, 5L, 5L, 5L, 5L, 5L, 5L, NA, 5L, 5L, 5L, NA, 5L, NA, NA, 
    5L, 5L, NA, NA, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L), median_secs = c(6.1902577, 
    6.1713217, 19.9907639, 19.7675571, 6.1000215, 6.1203305, 
    13.3964799, 7.0342887, 6.2597127, 6.2432609, 19.9681676, 
    20.1874484, 5.6448957, 5.6632873, 13.6564839, 5.5813086, 
    10.4640203, 10.8960488, 22.8774932, NA, 8.5460365, 12.2154671, 
    14.2776151, NA, 5.9763782, NA, NA, 5.6241829, 5.7766581, 
    NA, NA, 20.3630658, 10.6481448, 10.864006, 23.0229502, 23.5587724, 
    9.8157773, 13.6417876, 15.7693984, 9.0004465)), row.names = c(NA, 
-40L), class = c("data.frame", "data.table"))

```

```{r echo=FALSE}
library(data.table)
k <- 30 
setDT(dat)
pl(dat)
```

The example above is based on a no-frills equality join to allow comparison with
`merge.data.table` and `collapse::join()`, which have limited feature sets; see
the [Performance](https://trobx.github.io/fjoin/articles/fjoin-performance.html)
article for more detail. In the inner and left joins, `r pkg("fjoin")` and 
`merge.data.table` reflect a simple operation in `r pkg("data.table")`, and
directly inherit its excellent performance. But `r pkg("fjoin")` performs a bit
better than `merge.data.table` on the right and full joins. This is typical:
`r pkg("fjoin")`'s solutions for join types and additional options that are not
straightforward in native `r pkg("data.table")` have been developed with close
attention to performance.

# Notes for tidyverse users
`r pkg("fjoin")` is a drop-in alternative to `dplyr::*_join()` with 
faster large data performance and useful options that `dplyr::*_join()` doesn't
support (though the reverse is also true --- see below). Joins are infrequent
operations, and the package name is short, so you may not feel the need to
attach it:

```{r message=FALSE}
library(dplyr)
dfQ <- as_tibble(dfQ)

dfQ |>
  fjoin::fjoin_full(dfP,
                    on = "id",
                    select = c("item", "price", "quantity"),
                    order = "right",
                    indicate = TRUE) |>
  mutate(quantity = if_else(.join == 2, 0, quantity), 
         revenue = price * quantity)
```

`r pkg("fjoin")` is silent and permissive about cardinality: it currently has
no equivalent of `dplyr::*_join()`'s `relationship` validation. Nor does it yet
support rolling joins, which `r pkg("dplyr")` achieves elegantly via helper
functions in `join_by`. These features will be added. Nested/list‑column joins
(supported by `dplyr::*_join()`) are not planned.

The `r pkg("data.table")`-backed implementation of joins in `r pkg("dtplyr")`'s
is apparently no longer maintained and should be avoided. It only supports
equality joins, malfunctions with same-named columns on each side, and
silently ignores `r pkg("dplyr")`'s additional join arguments such as 
`na.matches` and `multiple` (as well as `relationship`). Use of
`r pkg("dtplyr")` also requires object conversions.

# Notes for `r pkg("sf")` users

TODO:

- attribute joins between `sf` objects are permitted (not in `r pkg("dplyr")`)
- `sfc`-class output columns always detected and refreshed regardless of whether
output is `sf` and whether they are the active geometry
- an example

# Notes for `r pkg("data.table")` users

`r pkg("fjoin")` automates joins that are challenging or laborious to write in
`r pkg("data.table")`, solving frustrations such as garbled join columns in
inequality joins and the lack of an effective `incomparables` argument, and 
providing other useful options. Even for simple joins, there is little reason
*not* to use it, since it is as fast or faster than `merge` and even native `r
pkg("data.table")` (see below).

It doesn't provide the full versatility of `r pkg("data.table")` joins, such as:

- computing on joined columns inside `j`, rather than simply selecting them
- joins by reference of the direct form `DT[i, on = <on> , v := i.v]`
- for now, rolling and overlap joins

You can nevertheless get a headstart by setting `do = FALSE`, copying the
console output, and editing the `j`-expression(s). Also, the assignment pattern
`DT[, v := fjoin_left(DT, i, on = <on>, select = "v")$v]` can be used to add new
join-based columns, and can usefully be combined with the `".join"` indicator.

```{r}
library(data.table)
dtQ <- as.data.table(dfQ)
dtP <- as.data.table(dfP)

dtP[, revenue := price * fjoin_left(
  dtP,
  dtQ,
  on = "id",
  select = c("quantity"),
  indicate = TRUE
  )[.join == 1L, quantity := 0]$quantity][]
```

The package actually began life as a tool to solve `r pkg("data.table")`'s
garbling of join columns in non-equi joins. It still does that:

```{r}
DT1 <- data.table(t=c(5,25,45))
DT2 <- data.table(t_start=c(1,21), t_end=c(10,30))
```

Range join with `r pkg("fjoin")`:
```{r}
dtjoin(DT2, DT1, on=c("t_start <= t", "t_end >= t"), show = TRUE)
```

Compare the default output from `r pkg("data.table")`:
```{r message=FALSE}
DT2[DT1, on=.(t_start <= t, t_end >= t)]
```

`r pkg("fjoin")` uses `data.frame()` coupled with `setDT()` instead of the more
usual `list()` in `j`. This is to sidestep an unnecessary deep copy of the
joined columns. Likewise, `r pkg("fjoin")` avoids deep copies on the way in and
out by "shallow-casting" inputs and outputs to and from `data.table`s as
necessary. Shallow conversion is not a safe general way of operating with
foreign objects in `r pkg("data.table")`, but it works for joins, because the
input vectors only need to be read, and the output vectors are guaranteed to be
unshared.

One consequence of this is that `r pkg("fjoin")` is more efficient than
`r pkg("data.table")` itself with standard interactive idioms. Consider the 
following (exaggerated) example, where the `i` expression is a `data.frame` and
we select columns in `j` with `list()` (aliased `.()`):

```{r message=FALSE}
n <- 1e3L; ncolDT <- 2L; ncolDF <- 10L
DT <- data.table(id = rep(1:n, each = 5L), matrix(runif(n * ncolDT), ncol = ncolDT))
DF <- data.frame(id = 1:n, matrix(runif(n * ncolDF), ncol = ncolDF))

bench::mark(
  iterations = 3,
  check      = TRUE,
  data.table = DT[DF, on = .(id), .(id, V1, V2, X1, X3, X5, X7, X9)],
  fjoin      = dtjoin(DT, DF, on = "id", select.i = c("X1", "X3", "X5", "X7", "X9"))
) |> summary() |> subset(select = c("expression", "n_itr", "median",  "mem_alloc"))
```

Here `r pkg("fjoin")` avoids a call to `as.data.table.data.frame` on the way in, 
and a call to `as.data.table.list` on the way out, both of which (currently)
always deep-copy. The `bench::mark` memory measurements
[exclude C-level allocations](https://rdatatable-community.github.io/The-Raft/posts/2025-07-14-memory_R-jangorecki/), but the difference between them reflects these R-level copies. This behaviour
will eventually change in `r pkg("data.table")`, at which point `r pkg("fjoin")`
will revert to the more familiar `j = list()` idiom.

# "Internals"

With `match.na = FALSE` (the default), `r pkg("fjoin")` inspects the data and,
if the join permits, chooses which table to omit `NA`-containing rows from using
a heuristic. The only thing that is not reflected in the `r pkg("data.table")`
code that `r pkg("fjoin")` produces are the input and output handling steps:

- shallow-casting non-`data.table` objects as `data.table`s
- adding tibble and/or `sf` class to outputs (and refreshing `sfc` columns
regardless of output class)
- on exit, dropping temporary columns (pointers) from `data.table` inputs used
as-is

These ensure that `r pkg("fjoin")` handles object classes with maximum
efficiency (no data copying) while leaving inputs intact.

In other respects, the package is a builder of `r pkg("data.table")` code to a
set of carefully developed solutions under a well thought-out API.
`r pkg("fjoin")` pays a lot of attention to doing this well and is also very
thoroughly tested.
