---
title: "Guide to fjoin"
format: html
vignette: >
  %\VignetteEngine{quarto::html}
  %\VignetteIndexEntry{Guide to fjoin}
  %\VignetteEncoding{UTF-8}
# knitr chunk options:
knitr:
  opts_chunk:
    collapse: false   # don't collapse code and output into one block
    out.width: "100%" # figs 100% width
# execution settings for Quarto:
execute:
  echo: true
  warning: false
  message: false
  cache: false # always render from scratch
---

<style>
/* 1. Code chunks: wrap long lines and remove scrollbars */
div.sourceCode {
  overflow-x: visible !important;
}
div.sourceCode pre,
pre.sourceCode {
  white-space: pre-wrap !important;
  word-break: break-word !important;
}
/* 2. Output chunks: Quarto-like styling */
.cell-output-stdout pre,
.cell-output-display pre,
.cell-output pre {
  background: white !important;
  border: none !important;
  padding: 0.4em 0 !important;
  white-space: pre-wrap !important;
  word-break: break-word !important;
}
/* 3. Output chunks: remove syntax highlighting, disable link styling (underline) and clicking */
.cell-output-stdout pre *,
.cell-output-display pre *,
.cell-output pre * {
  color: black !important;
  text-decoration: none !important;
  pointer-events: none;
}
/* 4. Reduce code and output font sizes and line spacing */
div.sourceCode pre,
pre.sourceCode,
.cell-output-stdout pre,
.cell-output-display pre,
.cell-output pre {
  font-size: 0.90em !important;
  line-height: 1.2em !important;
}
</style>

```{r setup, include=FALSE}
pkg <- function(x) {
  # style "pkgname" is defined in _pkgdown.yml 
  sprintf('<span class="pkgname">%s</span>', x)
}
```  
  
`r pkg("fjoin")` builds on `r pkg("data.table")` to provide fast, flexible joins
on any data frames. It slots into tidyverse pipelines and general workflows in a
single line, and provides NA-safe matching by default, on-the-fly column
selection, flexible row-order preservation, multiple-match handling on both
sides, and an indicator column for row origin.

# Installation

<!--
Stable release (CRAN):
```r
install.packages("fjoin")
```

-->
`r pkg("fjoin")` is currently in submission with CRAN.

Latest development version ([R-universe](https://trobx.r-universe.dev)):
```r
install.packages("fjoin",
  repos = c("https://trobx.r-universe.dev", "https://cloud.r-project.org"))
```

# Features

- **Scope:** Inner, left, right, full, semi-, and anti-joins with equality and
inequality conditions, plus cross joins.
- **Compatibility:** Accepts any mix of data frame-like objects or `list`s of
same-length vectors, and returns a plain `data.frame`, (grouped) tibble,
`data.table`, `sf`, or `sf`-tibble. Refreshes dynamic attributes like `groups`,
keys, `agr`, and `bbox` in the output.
- **Performance:** Uses case-specific `r pkg("data.table")` solutions specially
developed for high performance, and avoids data-copying at any stage for
memory-efficiency.
- **Transparency:** "Just works" for general users, but you can also view the
resulting `r pkg("data.table")` code instead of (or as well as) executing it, or
run “mock” joins without data that output template code.
- **Low dependency:** Depends only on `r pkg("data.table")`, a mature,
dependency-free package emphasising long-term stability.

`r pkg("fjoin")` provides several distinctive options and controls, including:

- **NA-safe joins by default:** Missing values are treated sensibly (no matches
on `NA` unless you actually ask for them with `match.na = TRUE`).
- **On-the-fly column selection:** Choose and order columns with `select` for
efficient one-liners.
- **Row origin tagging:** Add a Stata-style marker column showing each row’s
source with `indicate`.
- **Multiple-match handling on both sides:** Arguments `mult.x` and `mult.y`
specify how to handle multiple matches on either or both sides of the join,
including in semi- and anti-joins.
- **Flexible order:** Control which data frame's row order is preserved in the
result (without expensive sorting) using `order`.

# API

| `fjoin_*` functions<br>(`x`/`y` style) | `dtjoin_*` functions<br>(extended `DT[i]` style) |
|:-------|:--|
| `fjoin_inner()`, `fjoin_left()`, `fjoin_right()`,`fjoin_full()` | `dtjoin()` |
| `fjoin_left_semi` (alias `fjoin_semi`), `fjoin_right_semi()` | `dtjoin_semi()` |
| `fjoin_left_anti` (alias `fjoin_anti`), `fjoin_right_anti()` | `dtjoin_anti()` |
| `fjoin_cross()` | `dtjoin_cross()` |

The `fjoin_*` family consists of conventional left/right-style join functions. 
They are wrappers around the `dtjoin_*` functions (also exported), which write
the join code and handle execution, using an interface that generalises 
`r pkg("data.table")`'s `DT[i]` join syntax.

# Examples

```{r}
library(fjoin)
read_df <- function(x) data.table::fread(x, quote = "'", data.table = FALSE)
```

## Example 1: Basic use and options

Plain data frames joined by simple equality, using `fjoin_full()` for
illustration.
```{r}
dfP <- read_df("
id     item price other_cols
NA   apples    10        ...
 3  bananas    20        ...
 2 cherries    30        ...
 1    dates    40        ...
")

dfQ <- read_df("
id quantity       note  other_cols
 2        5         ''         ...
 1        6         ''         ...
 3        7         ''         ...
NA        8  'oranges'         ...
")
```

### (1) Basic syntax
```{r}
fjoin_full(dfQ, dfP, on = "id")
```
Notice that the default `match.na = FALSE` avoids an apples-and-oranges match
on missing values --- other R frameworks would join the last two rows.

### (2) Efficient selective joins in one line
Use `select` to restrict the join to particular columns and put them in desired
order:
```{r}
fjoin_full(dfQ, dfP, on = "id", select = c("item", "price", "quantity"))
```

This code is much easier to write and read than the equivalent operation in 
`r pkg("dplyr")`, which requires two calls before the join to avoid inflating
it with irrelevant columns, and a call after it to shuffle column order:
```{r eval = FALSE}
x <- dfQ |> select(id, quantity)
y <- dfP |> select(id, item, price)
full_join(x, y, join_by(id), na.matches = "never") |>
  select(id, item, price, quantity)
```

### (3) Indicator column for row origin
`1L`: left only, `2L`: right only, `3L`: joined from both. This extremely useful
feature has existed in Stata since 1984^[See p. 99 of the Stata 1 reference
manual, which has been linked to on Statalist and can be found by searching
online. Thanks to Nick Cox for this information.]

```{r}
fjoin_full(
  dfQ,
  dfP,
  on = "id",
  select = c("item", "price", "quantity"),
  indicate = TRUE
)
```

### (4) Switch row order from left-then-right to right-then-left
```{r}
fjoin_full(
  dfQ,
  dfP,
  on = "id",
  select = c("item", "price", "quantity"),
  indicate = TRUE,
  order = "right"
)
```

### (5) Display code instead of executing
For `r pkg("data.table")` users.
```{r}
fjoin_full(
  dfQ,
  dfP,
  on = "id",
  select = c("item", "price", "quantity"),
  indicate = TRUE,
  order = "right",
  do = FALSE
)
```

## Example 2: Reducing an inequality join from M:M to 1:1 with `mult.x` and `mult.y`

`r pkg("data.table")` (`mult`) and `r pkg("dplyr")` (`multiple`) can reduce the
cardinality on one side of the join from many (`"all"`) to one (`"first"` or
`"last"`). `r pkg("fjoin")` (`mult.x`, `mult.y`) will do this on either side of
the join, or on both sides at the same time. This example (using `fjoin_left()`)
shows an application to temporally ordered data frames of generic "events" and
"reactions".

```{r}
events <- read_df("
event_id event_ts
       1       10
       2       20
       3       40
")

reactions <- read_df("
reaction_id reaction_ts
          1          30
          2          50
          3          60
")
```

### (1) For each event, all subsequent reactions (M:M)
```{r}
fjoin_left(
  events,
  reactions,
  on = c("event_ts < reaction_ts")
)
```

### (2) For each event, the next reaction (1:M)
Equivalent to a one-way forward rolling join.
```{r}
fjoin_left(
  events,
  reactions,
  on = c("event_ts < reaction_ts"),
  mult.x = "first"
)
```

### (3) For each event, the next reaction, provided there was no intervening event (1:1)
Equivalent to a two-way rolling join (mutual forward/backward rolling matches).
```{r}
fjoin_left(
  events,
  reactions,
  on = c("event_ts < reaction_ts"),
  mult.x = "first",
  mult.y = "last"
)
```

## Example 3: Chain of calls from `fjoin_*` to `dtjoin_*` to `r pkg("data.table")`

A technical illustration showing:

- two calls to `fjoin_left()` (commented out), differing in the `order` argument
- the resulting calls to `dtjoin()`, with the addition of `show = TRUE`
- the generated `r pkg("data.table")` code and its output

```{r}
df_x <- data.frame(id_x = 1:3, row_x = paste0("x", 1:3))
df_y <- data.frame(id_y = rep(4:2, each = 2L), row_y = paste0("y", 1:6))
```

```{r}
# (1) fjoin_left(df_x, df_y, on = "id_x == id_y", mult.x = "first")
dtjoin(
  df_y,
  df_x,
  on = "id_y == id_x",
  mult = "first",
  i.home = TRUE,
  prefix = "R.",
  show = TRUE
)

# (2) fjoin_left(df_x, df_y, on = "id_x == id_y", mult.x = "first", order = "right")
dtjoin(
  df_x,
  df_y,
  on = "id_x == id_y",
  mult.DT = "first",
  nomatch = NULL,
  nomatch.DT = NA,
  prefix = "R.",
  show = TRUE
)
```

The difference in the `order` argument passed to `fjoin_*()` is reflected at
`dtjoin()` level in the identity of the tables passed to `.DT` and `.i`, the
values of the extended arguments `nomatch.DT` and `mult.DT` (counterparts to the
familiar `r pkg("data.table")` arguments `nomatch` and `mult` on the other side
of the join), and a compensating argument `i.home` which toggles the "home" and
"foreign" table for the purposes of column order and prefixing (as well as for
`indicate` and output class). `dtjoin()` in turn translates these specifications
into `r pkg("data.table")` code for execution. See the `dtjoin()` documentation
for full details of this extended `DT[i]`-style syntax.

# Performance

```{r echo=FALSE}
pal <- c(
  "#B8860B",
  "grey60",
  "#66C2A5",
  "#1B9E77"
)

factor_in_order <- function(x, ...) factor(x, levels = unique(x), ...)

pl <- function(x) {
  library(ggplot2)
  ggplot(x, aes(x = soln, y = ifelse(is.na(median_secs), 0, median_secs), fill = soln)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = sprintf("%.1f", median_secs)), vjust = -0.3, size = 3) +
    facet_grid(factor_in_order(style) ~ factor_in_order(xyargs), switch = "y") +
    scale_fill_manual(values = pal) +
    scale_y_continuous(limits = c(0, k)) +
    #theme_minimal() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      #strip.placement.y = "left",
      strip.text.y.left = element_text(angle = 0),
    ) +
    labs(
      x = NULL,
      # title = x[, unique(description)],
      subtitle = sprintf("median time in secs, %s runs per join",x[, unique(na.omit(N))])
      )
}

dat <- structure(list(style = c("Inner join", "Inner join", "Inner join", 
"Inner join", "Inner join", "Inner join", "Inner join", "Inner join", 
"Left join", "Left join", "Left join", "Left join", "Left join", 
"Left join", "Left join", "Left join", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by left)", 
"Right join (ordered by left)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Right join (ordered by right)", 
"Right join (ordered by right)", "Full join", "Full join", "Full join", 
"Full join", "Full join", "Full join", "Full join", "Full join"
), xyargs = c("x small, y big", "x small, y big", "x small, y big", 
"x small, y big", "x big, y small", "x big, y small", "x big, y small", 
"x big, y small", "x small, y big", "x small, y big", "x small, y big", 
"x small, y big", "x big, y small", "x big, y small", "x big, y small", 
"x big, y small", "x small, y big", "x small, y big", "x small, y big", 
"x small, y big", "x big, y small", "x big, y small", "x big, y small", 
"x big, y small", "x small, y big", "x small, y big", "x small, y big", 
"x small, y big", "x big, y small", "x big, y small", "x big, y small", 
"x big, y small", "x small, y big", "x small, y big", "x small, y big", 
"x small, y big", "x big, y small", "x big, y small", "x big, y small", 
"x big, y small"), soln = structure(c(1L, 2L, 3L, 4L, 1L, 2L, 3L, 
4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 
4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 4L, 1L, 2L, 3L, 
4L), levels = c("fjoin", "merge", "dplyr", "collapse"), class = "factor"), 
    description = c("No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered", 
    "No missing values, tables unordered", "No missing values, tables unordered"
    ), N = c(5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 
    5L, 5L, 5L, 5L, 5L, 5L, 5L, NA, 5L, 5L, 5L, NA, 5L, NA, NA, 
    5L, 5L, NA, NA, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L), median_secs = c(6.1902577, 
    6.1713217, 19.9907639, 19.7675571, 6.1000215, 6.1203305, 
    13.3964799, 7.0342887, 6.2597127, 6.2432609, 19.9681676, 
    20.1874484, 5.6448957, 5.6632873, 13.6564839, 5.5813086, 
    10.4640203, 10.8960488, 22.8774932, NA, 8.5460365, 12.2154671, 
    14.2776151, NA, 5.9763782, NA, NA, 5.6241829, 5.7766581, 
    NA, NA, 20.3630658, 10.6481448, 10.864006, 23.0229502, 23.5587724, 
    9.8157773, 13.6417876, 15.7693984, 9.0004465)), row.names = c(NA, 
-40L), class = c("data.frame", "data.table"))

```

```{r echo=FALSE}
library(data.table)
k <- 30 
setDT(dat)
pl(dat)
```

The benchmark above is based on a no-frills equality join to allow comparison with
`merge.data.table` and `collapse::join()`; see the [Performance](https://trobx.github.io/fjoin/articles/fjoin-performance.html)
article for more detail. In the inner and left joins, `r pkg("fjoin")` and 
`merge.data.table` reflect a simple operation in `r pkg("data.table")`, and
straightforwardly inherit its speed and robustness to the order of the tables.
But `r pkg("fjoin")` performs a bit better than `merge.data.table` on the right
and full joins. This is typical: `r pkg("fjoin")`'s solutions for join types and
additional options that are not straightforward in native `r pkg("data.table")`
have been developed with close attention to performance.

# Notes for `r pkg("tidyverse")` users
`r pkg("fjoin")` is a drop-in alternative to `dplyr::*_join()` with fast large
data performance and useful options that `dplyr::*_join()` lacks (though the
reverse is also true --- see below). Joins are fairly infrequent operations, and
the package name is short, so you may not feel the need to attach it:

```{r message=FALSE}
library(dplyr)
dfQ <- as_tibble(dfQ)

dfQ |>
  fjoin::fjoin_full(dfP,
    on = "id",
    select = c("item", "price", "quantity"),
    order = "right",
    indicate = TRUE
  ) |>
  mutate(
    quantity = if_else(.join == 2L, 0L, quantity),
    revenue  = price * quantity
  )
```

Please note that `r pkg("fjoin")`, for now, has no equivalent of
`dplyr::*_join()`'s `relationship` validation: it is silent and permissive about
cardinality. It also doesn't yet support rolling joins on unordered data,
which `r pkg("dplyr")` implements elegantly via a helper function in `join_by`,
or dedicated overlap joins (although these are easily written in terms of
inequalities). These features will be added.

The implementation of joins in the `r pkg("data.table")`-backed packages
`r pkg("dtplyr")` and `r pkg("tidytable")` needs maintenance. In both cases it
only supports equality joins, malfunctions in the presence of same-named
non-join columns on each side (try it), and silently ignores `r pkg("dplyr")`'s
additional join arguments such as `na.matches` and `multiple` (as well as
`relationship`). These packages are excellent tools that do much more than
just joins, but you should be aware of these limitations.

# Notes for `r pkg("sf")` users

```{r}
countries <- read_df("
country_id   country_name                          country_shape
         1     'Country A'  'POLYGON ((0 0, 1 0, 1 1, 0 1, 0 0))'     
         2     'Country B'  'POLYGON ((1 1, 2 1, 2 2, 1 2, 1 1))'    
         3     'Country C'  'POLYGON ((2 2, 3 2, 3 3, 2 3, 2 2))'    
") |> sf::st_as_sf(wkt = "country_shape", crs = 4326)

capitals <- read_df("
country_id  capital_name        capital_loc
         2       'City B'  'POINT (1.5 1.5)'    
         3       'City C'  'POINT (2.5 2.5)'    
         4       'City D'  'POINT (3.5 3.5)'    
") |> sf::st_as_sf(wkt = "capital_loc", crs = 4326)
```

`r pkg("fjoin")` smoothly accommodates joins involving `sf` data frames. In
particular, joins between two `sf` objects work as you would hope:

```{r}
fjoin_inner(countries, capitals, on = "country_id")
```

This is useful in workflows where you want to hold multiple geometries in the
same `sf` data frame. In `r pkg("dplyr")` such joins are prohibited:

```{r, message = TRUE}
try(dplyr::inner_join(countries, capitals, by = "country_id"))
```

In addition, `r pkg("fjoin")` always detects and refreshes `sfc`-class columns
in the join output, regardless of whether the inputs and output have `sf` class
(and whether those columns were/are active geometries). This avoids stale
bounding boxes and ensures that values at non-matching rows are converted from
`NULL` to a valid empty geometry. For example, here is the `sfc` column
`capital_loc` from the input on the right, after a left join in which the inputs
are plain data frames instead of `sf`s:

```{r, message=TRUE}
fjoin_left(as.data.frame(countries), as.data.frame(capitals), on = "country_id")$capital_loc
```

# Notes for `r pkg("data.table")` users

`r pkg("fjoin")` automates joins that are challenging or laborious to write in
`r pkg("data.table")`, while solving frustrations such as garbled join columns
in inequality joins and the lack of an effective `incomparables` argument, and 
providing other useful options. Even for very simple joins, there is no reason
*not* to use it, since it has negligible overhead and if anything will actually
slightly outperform `merge` (see above) and even native `r pkg("data.table")`
(see below).

That said, `r pkg("fjoin")` is not a comprehensive wrapper for 
`r pkg("data.table")`'s rich join functionality. Some things it cannot do are:

- computing on joined columns inside `j` (including `by = .EACHI` aggregations),
rather than simply selecting them
- joins by reference of the direct form `DT[i, on = <on> , v := i.v]`
- for now, dedicated rolling and overlap joins

You do have the option of setting `do = FALSE`, copying the console output, and
editing the `j`-expression(s). You can also "plonk" joined columns by reference
using the pattern `DT[, v := fjoin_left(DT, i, on = <on>, select = "v")$v]`,
which can often usefully be combined with the `".join"` indicator:

```{r}
library(data.table)
dtQ <- as.data.table(dfQ)
dtP <- as.data.table(dfP)

dtP[, revenue := 
      price * fjoin_left(
        dtP,
        dtQ,
        on = "id",
        select = c("quantity"),
        indicate = TRUE
      )[.join == 1L, quantity := 0L]$quantity][]
```

The package actually began life as a tool to solve `r pkg("data.table")`'s
garbling of join columns in non-equi joins. It still does that:

```{r}
dt1 <- data.table(t=c(5L,25L,45L))
dt2 <- data.table(t_start=c(1L,21L), t_end=c(10L,30L))
```

Here is a range join with `r pkg("fjoin")`:
```{r}
dtjoin(dt2, dt1, on=c("t_start <= t", "t_end >= t"), show = TRUE)
```

Compare the default output from `r pkg("data.table")`:
```{r message=FALSE}
dt2[dt1, on=.(t_start <= t, t_end >= t)]
```

Notice that `r pkg("fjoin")` uses `data.frame()` in `j` (coupled with an outer
`setDT()`) instead of the more usual `list()`, even when the required output is
a `data.table`. This is to sidestep an unnecessary deep copy of the joined
columns currently made by `r pkg("data.table")` in this case. Likewise,
`r pkg("fjoin")` avoids deep copies on the way in and out by "shallow-casting"
inputs and outputs to and from `data.table`s as necessary. Shallow conversion is
not a safe general way of operating with foreign objects in `r pkg("data.table")`,
but it works for joins, because the input vectors only need to be read, and the
output vectors are guaranteed to be unshared.

One consequence of this is that `r pkg("fjoin")` can be more efficient than
`r pkg("data.table")` itself with standard interactive idioms. Consider the 
following (exaggerated) example, where the `i` expression is a `data.frame` and
we select columns in `j` with `list()` (aliased `.()`):

```{r message=FALSE}
n <- 1e6L; ncol_dt <- 2L; ncol_df <- 10L
dt <- data.table(id = rep(1:n, each = 5L), matrix(runif(n * ncol_dt), ncol = ncol_dt))
df <- data.frame(id = 1:n, matrix(runif(n * ncol_df), ncol = ncol_df))

bench::mark(
  data.table = dt[df, on = .(id), .(id, V1, V2, X1, X3, X5, X7, X9)],
  fjoin      = dtjoin(dt, df, on = "id", select.i = c("X1", "X3", "X5", "X7", "X9")),
  iterations = 3,
  check      = TRUE
) |> summary() |> subset(select = c("expression", "n_itr", "median",  "mem_alloc"))
```

Here `r pkg("fjoin")` avoids a call to `as.data.table.data.frame` on the way in, 
and a call to `as.data.table.list` on the way out, both of which (currently)
always deep-copy. The `bench::mark` memory measurements
<a href="https://rdatatable-community.github.io/The-Raft/posts/2025-07-14-memory_R-jangorecki/" target="_blank">exclude C-level allocations</a>,
but the difference between them reflects these R-level copies. This behaviour
will eventually change in `r pkg("data.table")`, at which point `r pkg("fjoin")`
will revert to the more familiar `j = list()` idiom.

Finally, please note that `dtjoin()` differs from `r pkg("data.table")` in
its preservation of keys. In a `r pkg("data.table")` `DT[i]` join, the output
inherits the key of `DT` provided it happens to remain sorted on those
columns; this is consistent with `r pkg("data.table")`'s conception of
joins as a subsetting-like operation on `DT`, even though it is the
`i`-table that dictates the row order. With `dtjoin()`, the output always
inherits the key of `.i` (unless the non-joining rows of `.DT` are appended, in
which case the key is `NULL` since sortedness on `.i`'s key columns can no
longer be guaranteed). This design choice ensures intuitive behaviour of the
`fjoin_*()` functions: for example, `fjoin_left(x, y)` with the default 
`order = "left"` preserves `x`'s key, as a user would surely expect.

# "Internals"

With `match.na = FALSE` (the default), `r pkg("fjoin")` inspects the data and,
if the join permits, chooses which table to omit `NA`-containing rows from using
a heuristic. The only thing that is not reflected in the `r pkg("data.table")`
code that `r pkg("fjoin")` produces are the input and output handling steps:

- shallow-casting non-`data.table` inputs as `data.table`s (new objects
borrowing columns from the input)
- adding (grouped) tibble and/or `sf` class to `data.frame` outputs where
appropriate, including setting `groups` and `agr` attributes
- keying `data.table` outputs as described above
- refreshing `sfc` columns regardless of output class
- on exit, dropping temporary columns (pointers) from `data.table` inputs used
as-is

These steps ensure that `r pkg("fjoin")` handles object classes efficiently (no
data copying) while leaving inputs intact.

In other respects, the package is a constructor of `r pkg("data.table")` code to
a set of carefully thought-out solutions. `r pkg("fjoin")` pays a lot of
attention to doing this well and is also very thoroughly tested.
